---
title: "Linear regression with a multiple predictors I"
subtitle: "Lecture 14"
date: "2025-03-06"
format: revealjs
auto-stretch: false
---

<!-- begin: ae definition -->

```{r}
#| include: false
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
todays_ae <- "ae-11-modeling-penguins-multi"
```

<!-- end: ae definition -->

## While you wait...

::: appex
-   Go to your `ae` project in RStudio.

-   Make sure all of your changes up to this point are committed and pushed, i.e., there's nothing left in your Git pane.

-   Click Pull to get today's application exercise file: *`{r} paste0(todays_ae, ".qmd")`*.

-   Wait till the you're prompted to work on the application exercise during class before editing the file.
:::

# Linear regression with a categorical predictor

## From last time (with `penguins`)

::: question
A different researcher wants to look at body weight of penguins based on the island they were recorded on. How are the variables involved in this analysis different?
:::

. . .

-   outcome: body weight (numerical)

-   predictor: island (categorical)

## Visualize body weight vs. island 

::: question
Determine whether each of the following plot types would be an appropriate choice for visualizing the relationship between body weight and island of penguins.
:::

-   Scatterplot [❌]{.fragment}

-   Box plot [✅]{.fragment}

-   Violin plot [✅]{.fragment}

-   Density plot [✅]{.fragment}

-   Bar plot [❌]{.fragment}

-   Stacked bar plot [❌]{.fragment}

## Visualize 

::: task
Visualize the relationship between body weight and island of penguins. Also calculate the average body weight per island.
:::

```{r}
ggplot(penguins, aes(x = island, y = body_mass_g)) + 
  geom_point()
```

## Visualize 

::: task
Visualize the relationship between body weight and island of penguins. Also calculate the average body weight per island.
:::

```{r}
ggplot(penguins, aes(x = island, y = body_mass_g)) + 
  geom_boxplot()
```

## Visualize 

::: task
Visualize the relationship between body weight and island of penguins. Also calculate the average body weight per island.
:::

```{r}
ggplot(penguins, aes(x = island, y = body_mass_g)) + 
  geom_violin()
```

## Visualize 

::: task
Visualize the relationship between body weight and island of penguins. Also calculate the average body weight per island.
:::

```{r}
ggplot(penguins, aes(color = island, x = body_mass_g)) + 
  geom_density()
```

## Summarize, take 1

::: task
Visualize the relationship between body weight and island of penguins. Also calculate the average body weight per island.
:::

```{r}
penguins |>
  group_by(island) |>
  summarize(
    mean_bm = mean(body_mass_g)
  )
```

## Summarize, take 2

::: task
Visualize the relationship between body weight and island of penguins. Also calculate the average body weight per island.
:::

```{r}
penguins |>
  group_by(island) |>
  summarize(
    mean_bm = mean(body_mass_g, na.rm = TRUE)
  )
```

## Model - fit 

::: task
Fit a linear regression model predicting body weight from island and display the results. Why is Biscoe not on the output?
:::

```{r}
bm_island_fit <- linear_reg() |>
  fit(body_mass_g ~ island, data = penguins)

tidy(bm_island_fit)
```

## Model - interpret 

$$
\widehat{body~mass} = 4716 - 1003 \times islandDream - 1010 \times islandTorgersen
$$

-   Intercept: [Penguins from Biscoe island are expected to weigh, on average, 4,716 grams.]{.fragment}

-   Slope - islandDream: [Penguins from Dream island are expected to weigh, on average, 1,003 grams less than those from Biscoe island.]{.fragment}

-   Slope - islandTorgersen: [Penguins from Torgersen island are expected to weigh, on average, 1,010 grams less than those from Biscoe island.]{.fragment}

## Model - predict 

::: task
What is the predicted body weight of a penguin on Biscoe island? What are the estimated body weights of penguins on Dream and Torgersen islands? Where have we seen these values before?
:::

```{r}
new_penguins = tibble(
  island = c("Biscoe", "Dream", "Torgersen")
)

predict(bm_island_fit, new_data = new_penguins)
```

## Model - predict 

::: task
Calculate the predicted body weights of penguins on Biscoe, Dream, and Torgersen islands *by hand*.
:::

$$
\widehat{body~mass} = 4716 - 1003 \times islandDream - 1010 \times islandTorgersen
$$

. . .

-   Biscoe: $\widehat{body~mass} = 4716 - 1003 \times 0 - 1010 \times 0 = 4716$

. . .

-   Dream: $\widehat{body~mass} = 4716 - 1003 \times 1 - 1010 \times 0 = 3713$

. . .

-   Torgersen: $\widehat{body~mass} = 4716 - 1003 \times 0 - 1010 \times 1 = 3706$

## Models with categorical predictors

::: incremental
-   When the categorical predictor has many levels, they're encoded to **dummy variables**.

-   The first level of the categorical variable is the baseline level. In a model with one categorical predictor, the intercept is the predicted value of the outcome for the baseline level (x = 0).

-   Each slope coefficient describes the difference between the predicted value of the outcome for that level of the categorical variable compared to the baseline level.
:::
