---
title: "[What's it all about, Alfie?](https://youtu.be/2-he6twMEF0?feature=shared)"
subtitle: "Lecture 24"
date: "2025-4-22"
format: revealjs
auto-stretch: false
---

# Course admin

## While you wait

```{r}
#| label: load-packages
#| message: false
#| echo: false

library(tidyverse)
library(tidymodels)

todays_ae <- "ae-20-final-review"
```

-   Go to your `ae` project in RStudio.

-   Make sure all of your changes up to this point are committed and pushed, i.e., there's nothing left in your Git pane.

-   Click Pull to get today's application exercise file: *`{r} paste0(todays_ae, ".qmd")`*.

-   Wait until you're prompted to work on the application exercise during class before editing the file.

## Projects due tomorrow at 5PM

-   Written report in the `index` file;
-   Code chunks should *not* be displayed;
-   Five minute video presentation;
-   It needs to be obvious that all team members participated.

## Final exam next Tuesday 4/29 at 9AM

-   Roughly the same length as a midterm;
-   You have three hours to complete it;
-   You get both sides of an 8.5" x 11" sheet of notes;
-   All multiple choice;
-   Room assignments emailed to you;
-   20% of your final course grade;
-   Replaces a lower in-class midterm score;
-   Visit my OHs to view old midterm solutions.

## Style of questions

Things you've seen that will reappear:

-   Do you understand this picture?
-   Can you debug this incorrect code?
-   What is the correct statistical interpretation of this quantity?

A skill you will need:

> Can you picture the *intermediate output* of a data pipeline.

## From Midterm 1...

```{r}
#| echo: false
rm(n)
survey_counts <- tibble( 
  age = c(
    "18-49", "18-49", "18-49",
    "50+", "50+", "50+"
  ),
  opinion = c(
    "Remain Against The Law", "Be Made Legal", "Not Sure", 
    "Remain Against The Law", "Be Made Legal", "Not Sure"
  ),
  n_people = c(
    59, 292, (391 - (59 + 292)),
    67, 245, ((771 - 391) - (67 + 245))
  )
)
survey_counts
```

## From Midterm 1...

Why does this give an error?

```{r}
#| error: true
survey_counts |>
  group_by(age) |>
  summarise(
    age_sum = sum(n_people)
  ) |> 
  mutate(
    prop = n_people / sum(n_people)
  )
```

## How do the intermediate steps look?

```{r}
#| error: true
survey_counts 
```

## How do the intermediate steps look?

```{r}
#| error: true
survey_counts |>
  group_by(age)
```

## How do the intermediate steps look?

```{r}
#| error: true
survey_counts |>
  group_by(age) |>
  summarise(
    age_sum = sum(n_people)
  )
```

Indeed, `n_people` is not a column in the data frame that is being piped into `mutate`, hence the error.

## How do the intermediate steps look?

```{r}
#| error: true
survey_counts |>
  group_by(age) |>
  summarise(
    age_sum = sum(n_people)
  ) |> 
  mutate(
    prop = n_people / sum(n_people)
  )
```

## The full monty

-   *Preparation*: importing, joining, reshaping, mutating, etc;

-   *Exploration*: pictures and summaries;

-   *Modeling*: linear/logistic regression, model selection, etc.

-   *Inference*: interval estimation and hypothesis testing.

## `{r} todays_ae`

::: appex
-   Go to your ae project in RStudio.

-   If you haven't yet done so, make sure all of your changes up to this point are committed and pushed, i.e., there's nothing left in your Git pane.

-   If you haven't yet done so, click Pull to get today's application exercise file: *`{r} paste0(todays_ae, ".qmd")`*.

-   Work through the application exercise in class, and render, commit, and push your edits.
:::
